Is there a Criterion in Criterial Learning? \\ \vspace{.08in} Insights from Studying Feedback Delays

### 2025-10-17 write/response_to_reviews_GA.tex

\documentclass[12pt]{article}
% Language and font
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
% Bibliography with APA style
\usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{sample.bib}
% Layout and formatting
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\setstretch{1.0}
\usepackage{parskip} % disables indentation and adds vertical space between paragraphs
\setlength{\parindent}{0pt}
% Math and symbols
\usepackage{amsmath}
\usepackage{mathrsfs}
\usepackage{gensymb}
% Graphics and floats
\usepackage{graphicx}
\usepackage{float}
\usepackage{caption}
% Optional: nicer font
\usepackage{lmodern}
% Turn off section numbers
\setcounter{secnumdepth}{0}

\title{\begin{large}\textbf{Response to Reviews} \\ \vspace{.1in}
Is there a Criterion in Criterial Learning? \\ Insights from Studying Feedback Delays \\ \vspace{.1in}
\textit{Journal of Experimental Psychology: Learning, Memory, and Cognition} \\ \vspace{-.1in}
Manuscript number: XLM-2024-0436 \end{large}}
\author{
    Matthew J. Crossley, 
    Benjamin O. Pelzer,
    F. Gregory Ashby
}
\date{}

\begin{document}

\maketitle

We would like to thank the editor and reviewers for their
excellent suggestions and thoughtful comments. We worked
very hard to respond to all of these. As a result, we are
confident that the revised manuscript is significantly
stronger than our original submission. Please note that we
changed the title of the manuscript to more directly convey
the main point of our paper.

Our specific response to each reviewer comment is as
follows.

\section{Reviewer 1}
We were pleased that Reviewer 1 ``really enjoyed reading the
manuscript'' and ``highly recommend(ed) this paper to be
published after addressing a number of mostly minor
concerns.''

\subsection{Reviewer comment 1}
1. Exploration of model space: When reading the description
of the models it was not entirely clear to me whether the
three presented models explore the space of possible models
to a sufficient degree or whether there might be some models
missing from consideration that could change the overall
message.

\subsubsection{Author response}
This is an excellent and important point. In response, we
greatly expanded the set of models we explored, and we also
now classify these models more systematically. The previous
version examined 3 models, whereas the revision examines 12.
These 12 models are partitioned into two classes: one that
assumes a criterion is learned and one that assumes SR
learning with no criterion. The criterial-learning class
includes 8 different models that were created by factorially
combining three different binary assumptions: perceptual
drift (yes or no), criterial drift (yes or no), and
delay-sensitive updating (yes or no). The SR learning class
includes four models created by factorially combining two
binary assumptions: perceptual drift (yes or no) and
delay-sensitive feedback (yes or no).

\subsection{Reviewer comment 2}
2. Simulation results figures: There were some aspects of
the results figures for the simulation that I found
difficult to understand:

(a) Even after reading the full paper and thinking about it
for a while, I do not understand what panels B are supposed
to show. Does this only show the subset of the parameter
space in which one of the two relevant patterns was
observed? If not, shouldn't there be data points everywhere?

(b) I do not understand why in Figure 1, for example, panel
C shows some cases in which the predicted qualitative
pattern is either ``control impaired'' or ``other'', but there
is no corresponding bar in panel A.

(c) The bars in Figure 2A are smaller than in Figures 1A and
3A. Why?

\subsubsection{Author response}
With the new approach to model exploration, we have heavily
revised the simulation results figures.  We hope that the
new figures are clearer and address the reviewer's concerns.

\subsection{Reviewer comment 3}
3. Calling the simulation Experiment 1: I found it extremely
confusing that the simulation using parameter-space
partitioning was called ``Experiment 1''. If I read
``Experiment'' I expect to see some empirical data obtained
from participants and not solely a simulation. I suggest
renaming this section to something clearer. For example
``Simulation Study'' or ``Parameter-Space Partitioning
Across
Models''.

\subsubsection{Author response}
After thinking about this a bit, we agree that referring to
the computational modeling analyses as Experiment 1 was a
mistake. First, we agree with the reviewer that this
nomenclature is confusing, but we also came to realize that
the equations in this section might dissuade empirically
oriented readers from reading about our empirical results.
As a result, the revised manuscript renames the section
describing the computational modeling ``New Computational
Models of Criterial Learning'' and this more technical
section has been moved to later in the manuscript -- after
the two empirical experiments are described.

\subsection{Reviewer comment 4}
4. Outliers in Exp. 2:

- While I think the argument for removing outliers in Exp. 2
is fine, the actual criterion defining an outlier is not
actually given (which feels a bit ironic in a paper about
criteria).

\subsubsection{Author response}
Good point. We now say ``for each participant, we used the
best-fitting two component model to compute the likelihood
that their performance was best described by each component.
Participants whose data were estimated to most likely belong
to the low-performance distribution were classified as
outliers and excluded from further analysis.''

\subsection{Reviewer comment 5}
4. Outliers in Exp. 2:

- How do the results look with these outliers included? Does
the difference between condition vanish? Adding a footnote
or so should be enough.

- The fact that some participants did not even learn a
single criterion feels very surprising to me. Is this common
in such experiments? Please add a sentence embedding this
finding within your general experience running such
experiments.

\subsubsection{Author response}
When the outliers are included in the analysis, the
difference between conditions does indeed vanish. We have
added the following footnote: ``Including the excluded
outliers in these analyses eliminates the difference between
conditions.''

In our experience, it is not unusual for a small subset of
participants to perform very poorly in experiments of this
kind. This typically reflects a lack of engagement or
misunderstanding of the task instructions rather than an
inability to perform the task itself. We now state exactly
that in the revision.

\subsection{Reviewer comment 6}
Minor points:

- Does Eq. 13 show results for trial n or n + 1? The text
and equation are inconsistent.

\subsubsection{Author response}
We corrected this typo.

\subsection{Reviewer comment 7}
- The description regarding the randomisation of Experiment
2 on page 12 (paragraph before ``Procedure'') seems to be
inconsistent with Figure 4. Is the dimension picked at
random for each trial or always the same for certain problem
numbers (as implied by Figure 4)?

\subsubsection{Author response}
We agree that the description of the design of this
experiment (now Experiment 1) needed improvement. We revised
the paragraph Reviewer 1 refers to in this comment as
follows.

``Each participant practiced each of 14 one-dimensional
category-learning tasks, or problems, until they responded
correctly on 9 of the 10 previous trials, at which point the
problem changed. The 14 different category structures are
described in Figure 1. The relevant dimension was bar
thickness in problems 1 – 7 and bar angle in problems 8 –
14. Each problem included stimuli in two distinct clusters
that varied over a restricted range of the relevant
dimension. Critically, the optimal criterion value varied
from problem-to-problem with respect to its position within
the stimulus range. For example, for some problems the
optimal criterion was below the midpoint of the range and
for other problems it was above the midpoint.''

\subsection{Reviewer comment 8}
- In figure 4, the top says Problems ``1 thru 13'', but only
shows 7 problems.

\subsubsection{Author response}
We corrected this typo.

\vspace{.2in}

\section{Reviewer 2}

We were pleased that Reviewer 2 believed our  ``manuscript
reports a sophisticated approach to an important problem''
and that the reviewer ``especially like(d) the use of
mathematical models.''

\subsection{Reviewer comment 1}
I was concerned that the models seem to have multiple
varying features. It might be more useful to isolate
specific processes while holding other model elements
constant.

\subsubsection{Author response}
This is the same as Reviewer 1's first comment. As described
there, we strongly agree with this concern, and as a result,
we greatly expanded the number of models we examined, and we
reclassified these according to the fundamental assumption
of whether or not they assumed a criterion value is
specifically learned and stored trial-by-trial in memory.

\subsection{Reviewer comment 2}
I think it makes more sense to have a prediction simulation
section and Experiments 1 and 2 rather than to consider the
simulation the first experiment.

\subsection{Author response}
This is the same as Reviewer 1's third comment. As noted
there, we agree and as a result, we now refer to the two
empirical experiments as Experiment 1 and 2, and the
computational modeling section has been renamed to ``New
Computational Models of Criterial Learning'' and moved to
after the two experiments.

\subsection{Reviewer comment 3}
The data in Figure 6 can likely be fit much more closely by
an ExGaussian distribution than either the 1 or 2 component
Gaussian mixtures that are shown.

\subsubsection{Author response}
We thank the reviewer for this suggestion. We now estimate
both a single-component truncated ExGaussian and a
two-component truncated ExGaussian mixture via MLE with
proper normalization on the tasks bounded support $[9,
512]$, and compared them via AIC, BIC, and a parametric
bootstrap likelihood-ratio test (LRT).  The results from
moving from the mixture of Gaussians used in our orginal
submission to the truncated ExGaussian mixture used in the
revision does not change the qualitative pattern of our
results in any way. However, we agree that the ExGaussian
and ExGaussian mixture better capture the skewed and bounded
distribution of our trials-to-criterion data so we have
adopted it in the revision.

\subsection{Reviewer comment 4}
Replication of the criterial learning results from the
currently-labeled Experiment 2 seems like a good idea given
the fairly low sample sizes per condition.

\subsubsection{Author response}
We agree with the reviewer that a replication would be
valuable, but we believe the data and the expanded modeling
analysis currently presented represent a significant
contribution to the field.

\subsection{Reviewer comment 5}
Did stimuli remain on the screen across the feedback delay?
I don't think so. If not, what happens if the stimulus
reappears at the time of feedback? I could imagine this
would eliminate or dramatically reduce the negative effect
of feedback delay. If so, is that consistent with the
conclusions? That is, would the models that predict a
deleterious effect of feedback delay also predict that it
should be eliminated with ``reminder'' stimuli at the time
of feedback?

\subsubsection{Author response}
As shown in Figures 2 and 5, the stimulus was presented up
until a response was made. At this time the stimulus was
removed and replaced by a noise mask until feedback was
delivered. The question of how our results might have
changed if the stimulus were to reappear at the time of
feedback is both interesting and insightful. In the very
first study that investigated the effects of feedback delays
on category learning (Maddox, Ashby, \& Bohil, 2003,
JEP:LM$\&$C) we found that a short 2.5 s feedback delay
impaired information-integration category learning, but not
rule-based learning -- but only if a noise mask was
presented during the time between response and feedback. The
published 2003 article doesn't explore this issue, but we
hypothesized at the time that in the absence of the mask,
participants might be able to use eidetic imagery to keep
the relevant synapses active during the delay period. As a
result, we agree with the reviewer's prediction that
re-presenting the stimulus at the time of feedback would
eliminate the impaired learning that we observed. This
interesting point seems tangential to our main point,
however, and as a result, we did not address this point in
the revision.

\subsection{Reviewer comment 6}
For the currently-labeled Experiment 3, reporting
fail-to-reject results for t-tests is not a valid method for
establishing that feedback delay has no effect. This is
especially a concern given the low participant count per
condition and the fact that empirical effects go in the
``right'' direction (lower performance for delayed
feedback).

\subsubsection{Author response}
This is a completely valid point. As in almost all
psychological experiments, our Experiment 2 lacks the
statistical power to make strong inferences about the null
results we obtained there. We now explicitly acknowledge
this point in paragraph 2 of the section entitled
``Discussion of Experiments 1 and 2.'' Also, rather than try
to make a statistical argument in support of our null
results, we now use a converging operations approach by
appealing to the literature. The new relevant paragraph is
as follows.

``Whereas the results of Experiment 1 seem clear, the
Experiment 2 results are weaker since they suggest a null
result. Although insufficient power makes it impossible to
rule out the possibility that Experiment 2 missed some small
effect of feedback delay, the null results found there are
consistent with a wide variety of other evidence. In
particular, a variety of other studies have reported that
feedback delays of up to 10 s do not significantly slow
one-dimensional rule-based learning (Dunn et al., 2012; Ell
et al., 2009; Maddox et al., 2003; Maddox \& Ing, 2005;
Worthy et al., 2013). In addition, many studies have
reported evidence that one-dimensional rule-based learning
recruits working memory and executive attention, but not
procedural learning (for reviews, see, e.g., Ashby \&
Valentin, 2017; Ashby et al., 2020). So the results of
Experiment 2 are consistent with all of these studies. On
the other hand, several of these feedback-delay studies did
not use binary-valued stimulus dimensions, so some criterial
learning was required. If so, then why did they all report
no effect of feedback delay? It is important to note that
criterial learning was not the focus of any of these
studies, and as a result, in all of these previous studies
the optimal criterion was always set exactly midway between
the category prototypes, which makes criterial learning
trivial. For example, under these conditions, criterial
learning might not even require feedback. The unsupervised
category-learning experiments reported by Ashby et al.
(1999) provide strong support for this because all of their
rule-based participants learned the correct criterion (which
was midway between the category means), even though the task
was completely unsupervised.''

\subsection{Reviewer comment 7}
I recommend running a bigger replication study that combines
both tasks to demonstrate the claimed interaction (feedback
delay impairs performance only for the criterion learning
task). The replication study should be preregistered with a
sample size justification based on the interaction effect
defined by the across-experiment comparison in the current
data. Adding a ``stimulus reminder at feedback'' condition
to this new study might be helpful in clarifying the
mechanisms producing the delay effect.

\subsubsection{Author response}
We agree that a replication / extension study would
strengthen our manuscript -- as it would for almost all
empirical papers. We chose not to pursue this suggestion,
however.  Even so, we believe we did significantly
strengthen our manuscript by greatly expanding the class of
computational models that we investigated, and by better
situating our results within the existing literature.

\section*{Action Editor}

\subsection{AE comment 1}
The final paragraph of the introduction summarizes the
experiments. This seems more appropriate in the abstract,
given that no information about the experiments are
presented in the abstract.

\subsubsection{Author response}
Thanks for this suggestion. The revised abstract now has
more detail about the experiments.

\subsection{AE comment 2}
Equation 1 needs some explanation, as this implies that the
decision Rn is made immediately on stimulus presentation. In
other words, Rn is not an outcome of some time-extended
cognitive process. Am I misunderstanding equation 1?

\subsubsection{Author response}
This is a valid and insightful point. Equation 1 does assume
that the response is made immediately upon stimulus
presentation. Because we are not trying to account for
response time, we decided to not explicitly model the
time-extended cognitive processes that govern evidence
accumulation. We now state this explicitly in the revision.
Specifically, immediately after equation 1, we added the
following text:

``Note that this equation indicates that the response is
made immediately upon stimulus presentation. We therefore
are not explicitly modeling the time-extended cognitive
processes that govern evidence accumulation and the
corresponding response time.''

\subsection{AE comment 3}
What was the reason behind the choice to have equation 2
only be sensitive to negative feedback?

\subsubsection{Author response}
The rationale behind the choice to have equation 2 only be
sensitive to negative feedback is now stated explicitly in
the first two sentences of the paragraph that includes
equation 2. Those sentences read as follows.

``If positive feedback is received, then the value of the
criterion remains unchanged for the next trial (except for
drift – see equation 3). The rationale here is that if the
response is correct, then the observer has effectively
gained zero information about how their criterion should be
modified.''

\subsection{AE comment 4}
How is ``optimal'' defined in the current case, given that
response time does not factor in the update rule?

\subsubsection{Author response}
By ``optimal'', we mean the value of the criterion that
maximizes accuracy. We now say this explicitly in the
revision. This is now explained in the last sentence of the
paragraph that includes equation 2, which reads as follows.

``This updating rule will gradually converge on the optimal
criterion value (i.e., the value of the criterion that
maximizes accuracy).''

\subsection{AE comment 5}
The use of parameter space partitioning is sensible given
the many differences among the three models. However, PSP
operates on non-stochastic models. In addition, the way you
have approached PSP is not how the original PSP is
implemented (i.e., using MCMC). This should be mentioned
explicitly to avoid confusion.

\subsubsection{Author response}
Thanks for raising this important point. We have revised the
manuscript to make this point clear by adding the following
text  to the beginning of the third paragraph of the section
entitled ``Simulation Results.''

``Classical PSP is defined for non-stochastic models and is
often implemented by exploring the parameter space via a
Markov chain Monte Carlo algorithm. In contrast, we used
brute force search over a grid of plausible parameter
values. Because our models are stochastic, at each sampled
set of parameter values we generated multiple simulations
and classified the results into one of a small set of
qualitative outcome patterns.''

\subsection{AE comment 6}
The three-dimensional plots in figures 1B, 2B, and 3B do not
come out right on 2D. You could improve these by having 2D
projections on the planes. Better still, three boxplots
would be more appropriate as there is no additional
information by using 3D plots.

\subsubsection{Author response}
This was a good suggestion. We revised the two figures (8
and 9) that describe the simulation results accordingly --
i.e., both figures are 2D and include boxplots.

\subsection{AE comment 7}
Although not a focus in this manuscript, I wondered whether
the models are able to account for finer-grained data
patterns such as complete RT distributions and
speed-accuracy tradeoff functions.

\subsubsection{Author response}
As described in our response to AE comment 2, the models
make no attempt to account for response time (RT). The
models could all be generalized to make RT predictions, but
any such generalization would require adding extra
assumptions about the time course of the decision-making
process. The focus of our research was on \textit{what} is
updated trial-by-trial (i.e., a criterion vs SR
associations), rather than on the dynamics of the updating
process. For this reason, we chose a theoretically minimal
landscape. Specifically, we were concerned that adding any
extra assumptions that were not directly related to our
primary goal would weaken the inferences possible from our
model simulations. For example, it would be difficult to
rule out the possibility that any qualitative result we
established was a consequence of one of these added
assumptions, rather than because of a difference in what was
being updated on each trial. Even so, we agree that
extending the models proposed here to make RT predictions
could be a fruitful avenue for future work.

\subsection{AE comment 8}
The experiments are numbered 2 and 3, instead of 1 and 2.
The models were following the methods of experiment 1, which
is missing. Experiment 1 is also mentioned in the discussion
separately with experiment 2. It looks like experiment 1 has
been deleted from this manuscript, but the model simulations
and discussion still refer to it.

### 2025-10-17 write/main_revision.tex
updated analysis et al for revision

\title{Criterial Learning and Feedback Delay: Insights from
Computational Models and Behavioral Experiments}
\shorttitle{Computational Criterion Learning}

@@

\abstract{
    The notion of a response criterion is ubiquitous in
    psychology, yet its cognitive and neural underpinnings
    remain poorly understood. To address this shortcoming,
    three computational models that capture different
    hypotheses about criterial learning were developed and
    tested. The time-dependent drift model assumes the
    criterion is stored in working memory and that its value
    drifts over time. The delay-sensitive learning model
    assumes that the magnitude of criterial learning is
    temporally discounted by feedback delay. The
    reinforcement-learning model assumes that criterial
    learning emerges from stimulus-response association
    learning without an explicit representation of the
    criterion, with learning rate also temporally discounted
    by feedback delay. The performance of these models was
    investigated under varying feedback delay and intertrial
    interval (ITI) durations. The time-dependent drift model
    predicted that long ITIs and feedback delays both impair
    criterial learning. In contrast, the delay-sensitive and
    reinforcement-learning models predicted impairments only
    with feedback delays.  Two behavioral experiments tested
    these predictions. In Experiment 1, participants were
    explicitly instructed about the relevant dimension to
    isolate criterial learning, and short feedback delays
    impaired learning relative to immediate feedback. In
    Experiment 2, using binary-feature stimuli without
    instruction (isolating rule selection/switching),
    neither longer ITIs nor delayed feedback affected
    performance. These results support the delay-sensitive
    and reinforcement-learning models, and suggest that even
    in tasks that appear to rely on explicit, rule-based
    reasoning, criterial learning may have strong
    associative underpinnings.
}

@@

This article addresses this shortcoming through a
combination of computational modeling and empirical data
collection. Specifically, we develop and test three
different computational models that make qualitatively
different assumptions about how the criterion is learned.
The models differ in the role they assign to working memory
and in whether they treat the response criterion as a
fundamental psychological construct, or instead assume that
behavior is driven purely by stimulus-response associations
without any criterion guiding responses. These models are
then tested in two behavioral experiments. The modeling and
empirical focus are on how feedback delays and the length of
the intertrial interval (ITI) affect criterial learning. All
criterial-learning models assume that updating (i.e.,
learning) of the criterion occurs during the time interval
between feedback presentation and the stimulus presentation
that defines the onset of the next trial. So feedback delay
and the length of the ITI are the independent variables that
most clearly differentiate the conflicting predictions of
criterial-learning models. Furthermore, feedback delays are
known to impair some forms of learning (i.e., procedural)
much more than others (i.e., learning that relies on
declarative memory), so feedback-delay manipulations offer a
powerful method of disambiguating the nature of criterial
learning \parencite{ell2009critrial, MaddoxAshbyBohil2003,
MaddoxIng2005}.

Although the models could be tested in any task that depends
on criterial learning, the two experiments we describe used
a one-dimensional category-learning task. In such tasks,
stimuli vary across trials on two or more dimensions -- one
relevant to categorization and one or more that are
irrelevant. The observer's goal is to identify the relevant
dimension and learn the response criterion that maximizes
accuracy. This task has been used in hundreds of studies,
and all current models of performance in this task emphasize
the role of criterial learning.

In the first behavioral experiment, participants were
explicitly instructed about the relevant dimension, thereby
isolating criterial learning from rule selection and
switching processes. The results showed that short feedback
delays impaired learning compared to immediate feedback.
The second behavioural experiment used stimuli composed of
binary features -- where no criterial learning is needed --
and did not instruct participants about the relevant
dimension. Thus, this experiment isolated rule selection and
switching from criterial learning. The results found that
increasing feedback delay or ITI did not affect performance.
These findings suggest that feedback delays impact criterial
learning but not the discovery of the relevant stimulus
dimension.  More broadly, they indicate that criterial
learning may recruit procedural learning, even in tasks that
seem to rely on explicit, rule-based processes.

\section{New Computational Models of Criterial Learning}
We developed computational models of criterial learning
within two broad architectures and examined their
sensitivity to ITI duration and feedback delay. The
architectures differ in whether they assume that the
response criterion is explicitly represented in memory. In
the first architecture, the criterion is stored in memory,
and responses are generated by comparing the current percept
with this stored referent. The criterion is updated after
errors using a gradient-descent rule. Models in this class
can exhibit sensitivity to ITI and feedback delay in three
distinct ways: (1) perceptual drift, where stimulus
representations change over time, (2) criterial drift, where
the stored criterion shifts over time, and (3)
delay-sensitive updating, where the magnitude of criterion
updates decreases with longer feedback delays. In the second
architecture, no explicit criterion is stored.  Instead,
reinforcement learning forms direct stimulus--response
associations that drive performance.  Sensitivity to ITI and
feedback delay arises from either perceptual drift or
delay-sensitive updating, but criterial drift is not
possible since no criterion is represented.

\subsection{Models that assume a stored criterion}
These models assume that the observer constructs a response
criterion, maintains it in working memory, and makes
decisions by comparing the percept to this criterion. They
may further posit that both the stored criterion and the
perceived stimulus value drift over time, with the extent of
drift increasing both within a trial and across consecutive
trials. In addition, they may assume that the learning rate
for updating the criterion decreases as the feedback delay
increases.

Let $x_n(t)$ and $c_n(t)$ denote the values of the percept
and the criterion, respectively, on trial $n$ at time $t$.
Then the decision rule on trial $n$ is:
\begin{equation}
  \text{Respond } R_n =
  \begin{cases}
    \text{A}, & \text{if ~} x_n(\text{T}_\text{S}) \leq c_n(\text{T}_\text{S})  \\
    \text{B}, & \text{if ~} x_n(\text{T}_\text{S}) > c_n(\text{T}_\text{S}).
  \end{cases}
  \label{eq:DR}
\end{equation}
Note that this equation indicates that the response is made
immediately upon stimulus presentation. We therefore are not
explicitly modelling the time-extended cognitive processes
that govern evidence accumulation and the corresponding
response time.

If positive feedback is received, then the value of the
criterion remains unchanged for the next trial (except for
drift -- see Equation \ref{eq:criterion}).  The rationale
here is that if the response is correct, then the observer
has effectively gained zero information about how their
criterion should be modified.  If negative feedback is
received, then the criterion is modified according to the
standard model
\parencite{SuttonBarto1998}:
\begin{equation}
c_n(\text{T}_\text{F}+ \Delta_t) = c_n(\text{T}_\text{F}) + \frac{\alpha}{t_{FD}} [x_n(\text{T}_\text{F}) - c_n(\text{T}_\text{F})],
\label{RL}
\end{equation}
where $\alpha$ is a learning-rate parameter, and $\Delta_t$
is the time it takes to complete the updating.  It is
straightforward to show that the iterative Equation \ref{RL}
is equivalent to computing a weighted mean (weighted by
recency) of the values of all percepts that occur on error
trials \parencite[e.g.,][]{Ashby2017}. This updating rule
will gradually converge on the optimal criterion value
(i.e., the true criterion value the observer is trying to
learn).

The term $\frac{\alpha}{t_{FD}}$ captures the notion that
the magnitude of the update decreases as the feedback delay
increases.  As a result, when feedback is delayed, the
system becomes less responsive to errors, leading to slower
learning compared to immediate feedback conditions. However,
since this model is consistent with the idea that criterial
learning relies on working memory -- and the available
evidence suggests that logical reasoning and working memory
are unaffected by feedback delays of several seconds
\parencite[e.g., in one-dimensional rule-based category
learning tasks;][]{ell2009critrial, MaddoxAshbyBohil2003,
MaddoxIng2005} -- the learning process described in Equation
\ref{RL} likewise be unaffected by feedback delay. In this
case, the term $\frac{\alpha}{t_{FD}}$ is replaced by
$\alpha$, making the learning rate independent of feedback
delay.

This model further assumes that both the stimulus and the
criterion representations may drift randomly throughout the
duration of time they are maintained in working memory. The
representation of the criterion must always be maintained in
working memory, whereas drift in the stimulus representation
affects performance up until the feedback is presented, but
not afterwards. We modeled the drift in both the criterion
and the percept by adding white noise to their initial
values. Specifically, we assumed that for all
$t>\text{T}_\text{S}$
\begin{equation}
  c_n(t) = c_n(\text{T}_\text{S}) + \eta_c \epsilon(t),
  \label{eq:criterion}
\end{equation}
and
\begin{equation}
  x_n(t) = x_n(\text{T}_\text{S}) + \eta_x \epsilon(t),
  \label{eq:percept}
\end{equation}
where $\epsilon(t)$ is white noise and $\eta_c$ and $\eta_x$
are parameters that determine the amount of drift over time.
Note that these parameters are permitted to be equal to zero
which is equivalent to assuming that the criterion and/or
the percept do not drift over time at all.

This model predicts that at the time of feedback, the
response criterion $c_n(\text{T}_\text{F})$ will be normally
distributed with mean $c_n(\text{T}_\text{S})$ and variance
$\text{t}_\text{FD} \eta_c^2$. Similarly, at the time when
the stimulus that defines the next trial is presented, the
criterion $c_n(\text{T}_{\text{S}^+}) =
c_{n+1}(\text{T}_\text{S})$ is normally distributed with
mean $c_n(\text{T}_\text{S})$ and variance
$(\text{t}_\text{FD}+\text{t}_\text{ITI}) \eta_c^2$. The
predictions for the percept are similar. Specifically, at
the time when feedback is presented, the percept
$x_n(\text{T}_\text{F})$ is normally distributed with mean
$x_n(\text{T}_\text{S})$ and variance
$(\text{T}_\text{F}-\text{T}_\text{S}) \eta_x^2$.

\subsection{Reinforcement-learning models}
The reinforcement-learning models gradually associate
responses with stimuli, and therefore do not include an
explicit representation of the response criterion. The
models are based on an actor-critic architecture and use
reinforcement learning to form stimulus-response
associations \parencite{SuttonBarto1998}. These models
assumes that the perceptual representation of each stimulus
is defined by the pattern of activation across 25 sensory
units that are characterized by overlapping tuning curves.
Each unit is maximally excited by one specific stimulus,
which we call the unit's preferred stimulus. Specifically,
the activation of the $i^\text{th}$ sensory unit on trial
$n$ is given by
\begin{equation}
  A_i(n) = \text{exp} \left( \frac{-\left[d_{i,S_n}\right]^2}{\sigma^2} \right)
  \label{eq:DSL_activation}
\end{equation}
where $d_{i,S_n}$ is the (Euclidean) distance between the preferred
stimulus value of the $i^\text{th}$ sensory unit and the
value of the stimulus that was presented on trial $n$, and
$\sigma$ is a constant that increases with perceptual noise.

>>>>

\title{Is there a Criterion in Criterial Learning? \\ \vspace{.08in}
Insights from Studying Feedback Delays
}
\shorttitle{Criterial Learning}

@@

\abstract{
     The notion of a response criterion is ubiquitous in
     psychology, yet its cognitive and neural underpinnings
     remain poorly understood. Two experiments and extensive
     computational modeling were used to test between two
     strikingly different interpretations of the criterion.
     The traditional account is that decisions are made by
     comparing the stimulus value to a stored value of the
     criterion. A conceptually different interpretation is
     that learning instead is a process of associating
     responses with stimuli, and that the criterion is
     simply the hypothetical value that separates stimuli
     associated with contrasting responses. The experiments
     and modeling tested between these two interpretations
     by contrasting the effects on criterial learning of
     feedback delays versus increases in the duration of the
     intertrial interval in a one-dimensional
     category-learning task. The empirical results strongly
     suggested that human criterial learning is sensitive to
     feedback delay but not to the duration of the
     intertrial interval. The computational modeling showed
     that these results are compatible with a
     stimulus-response learning account, and incompatible
     with all versions of the stored-criterion account,
     except for the subset of these models that explicitly
     assume the criterial updating process is sensitive to
     feedback delay.
}

@@

This article addresses this shortcoming through a
combination of empirical data collection and computational
modeling. Specifically, our goal is to test between two
general possibilities. One is that the criterion is learned
in the sense that it is updated trial-by-trial and that its
current value is stored in memory. This is the hypothesis
implicitly assumed by most decision-making models (e.g.,
signal-detection theory). The second possibility is that the
criterion has no psychological meaning and learning instead
is of stimulus-response (SR) associations. According to this
account, the criterion is simply the sensory (or cognitive)
value that separates percepts associated with the
contrasting responses. For example, there is evidence that
procedural learning works in this way
\parencite{AshbyWaldron1999}.

Two independent variables that seem especially likely to
discriminate between these two alternatives are the duration
of the intertrial interval (ITI) and the delay between the
response and the feedback. If the value of a criterion is
updated following feedback and then held in memory until
needed again, then the longer it must be held in memory, the
more time there is for its value to drift. In other words,
there should be more criterial drift with long ITIs than
with short ITIs, and as a result, increasing the ITI should
impair performance. In contrast, if the criterion has no
psychological meaning and instead learning is of SR
associations, then learning should be impaired by increases
in feedback delay. The idea is that positive feedback
strengthens and negative feedback weakens recently active
synapses that associate a response with the presented
stimulus. As a result, increasing the feedback delay will
weaken the trace of recently active synapses, and therefore
reduce this type of SR learning. This model is supported by
many previous reports that feedback delays as short as 3 s
impair procedural learning much more than  declarative
learning that depends on executive attention and working
memory  \parencite{DunnEtAl2012, ell2009critrial,
MaddoxAshbyBohil2003, MaddoxIng2005,  Worthyetal2013}.

Our general approach is as follows. First, we establish an
empirical database by describing the results of two
experiments that report the results of manipulating ITI and
feedback delay on criterial learning. Although a variety of
different tasks could be used to study criterial learning,
the two experiments we describe used a one-dimensional
category-learning task. In such tasks, stimuli vary across
trials on two or more dimensions -- one relevant to
categorization and one or more that are irrelevant. The
observer's goal is to identify the relevant dimension and
learn the response criterion that maximizes accuracy. This
task has been used in hundreds of studies, and all current
models of performance in this task emphasize the role of
criterial learning.

In the first experiment, participants were explicitly
instructed about the relevant dimension, thereby isolating
criterial learning from rule selection and switching
processes. The results showed that increasing the feedback
delay impaired learning but increasing the ITI did not.
Experiment 2 used stimuli that varied on six binary-valued
dimensions and participants were not told which of these
dimensions was relevant to the categorization decision. The
observer's task was to discover the single relevant
dimension, but after this dimension was identified, no
criterial learning was required since there were two
categories and the stimuli had exactly two values on the
relevant dimension. Thus, Experiment 2 isolated rule
selection and switching from criterial learning. The results
showed that performance was unaffected either by increases
in feedback delay or ITI. These findings therefore suggest
that feedback delays impact criterial learning but not the
discovery of the relevant stimulus dimension.

The results of Experiments 1 and 2 seem to suggest that
criterial learning is a form of procedural learning, even in
tasks that seem to rely on explicit, rule-based processes,
and therefore that the criterion is an emergent construct
with no psychological meaning. Before accepting this
conclusion however, it is important to ask whether some type
of criterion-learning model can account for these results.
To investigate this question, we developed a variety of
computational models of criterial learning of two different
types. Eight different models assumed the criterion is
learned and updated trial-by-trial. These were created by
factorially combining three different binary assumptions:
the percept drifts randomly over time (yes or no), the
criterion drifts randomly over time (yes or no), and the
updating of the criterion is sensitive to the feedback delay
(yes or no). Four different models assumed that learning is
of SR associations and that the criterion therefore has no
psychological meaning. These were created by factorially
combining two binary assumptions: the percept drifts
randomly over time (yes or no) and SR strengthening is
sensitive to feedback delay (yes or no). Next, using a
technique called parameter-space partitioning
\parencite{pitt2006global}, we investigated whether each
model could account for the qualitative results of
Experiments 1 and 2. The results of these analyses
reinforced the conclusion that criterial learning may have
strong associative underpinnings, even in tasks that appear
to rely on explicit, rule-based reasoning,

\section{Experiment 1}
Experiment 1 used a one-dimensional category-learning task
to investigate how feedback delay and ITI affect criterial
learning. The stimuli in this experiment were circular
sine-wave gratings that varied across trials in bar width
and bar orientation. These stimuli were divided into two
categories according to their value on one of the two
dimensions. In other words, the optimal strategy was to set
a response criterion on the single relevant dimension, and
then choose a categorization response based on whether the
value of the presented stimulus on this dimension was larger
or smaller than the criterion value.  The experiment
isolated criterial learning by (1) explicitly instructing
participants about the relevant stimulus dimension and rule
structure (e.g., thick bars are ``A'', thin bars are ``B''),
and (2) eliminating all variability along the irrelevant
stimulus dimension. In other words, the instructions
identified the relevant stimulus dimension, and as a result,
the only learning required was criterial learning.

Each participant practiced each of 14 one-dimensional
category-learning tasks, or problems, until they responded
correctly on 9 of the 10 previous trials, at which point the
problem changed. The 14 different category structures are
described in Figure \ref{fig_design_exp_1_space}. The
relevant dimension was bar thickness in problems 1 -- 7 and
bar angle in problems 8 -- 14. Each problem included stimuli
in two distinct clusters that varied over a restricted range
of the relevant dimension. Critically, the optimal criterion
value varied from problem-to-problem with respect to its
position within the stimulus range. For example, for some
problems the optimal criterion was below the midpoint of the
range and for other problems it was above the midpoint.
Feedback delay and ITI were varied across three conditions.
In the Delayed-Feedback condition, feedback was delayed
after the observer responded (by 3.5 s) and the ITI was
short (500 ms). In the Long-ITI condition, the feedback
delay was short (500 ms) and the ITI was long (3.5 s).
Finally, in the Control condition, feedback delay and the
ITI were both 500 ms.
